---
title: "Homework 2"
author: "Ethan Jantz"
date: "4/10/2021"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)

library(tidyverse)

pbe <- read_csv(here::here("HW 3", "cavax.csv"))
concarry <- read_csv(here::here("HW 3", "concealed_carry.csv"))
```

# PART ONE

## QUESTION 1

A. In how many schools is the percentage of students exempt for medical reasons (`med_exempt`) greater than the percentage exempt for religious reasons (`rel_exempt`)? Of this set of schools, what percent are public schools?

```{r}
pbe %>%
  filter(med_exempt > rel_exempt) %>% # 518 Schools
  filter(type == "PUBLIC") # 448 public schools

print(round(448/518, 2)) # 86%
```

B. Which county, when averaging across all schools in that county, has the highest average percentage of exempt students (`exempt`)?

```{r}
pbe %>%
  group_by(county) %>%
  summarize(exempt_avg = mean(exempt)) %>%
  arrange(desc(exempt_avg)) # Nevada county
```

C. Create a bar chart that shows for private and public schools (`type`) the percent of students exempt from providing vaccination records (`exempt`). 

```{r}
pbe %>%
  group_by(type) %>%
  summarize(pct_exempt = mean(exempt) / 100) %>%
  ggplot(aes(x = pct_exempt, y = type)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent) + 
  geom_text(aes(label = scales::percent(pct_exempt)), hjust = 3)
```

## QUESTION 2

Estimate a model predicting `exempt` by district `type` and `enrollment.` Treat `exempt` as a continuous variable. Interpret the intercept and coefficients. What is the predicted `exempt` percentage for a public school with 100 students in kindergarten? What is the predicted `exempt` percentage for a private school with 80 students in kindergarten?

```{r}
pbe_lm <- lm(exempt ~ type + enrollment, data = pbe)

summary(pbe_lm)
```

## QUESTION 3

Test whether the assumption of homoskedaticity has been met. Discuss results. Calculate the VIF for each variable. Should we be concerned with multicolinearity?

```{r}
car::ncvTest(pbe_lm)

plot(pbe_lm, 3)
```

```{r}
car::vif(pbe_lm)
```

## QUESTION 4

Recenter the variable `enrollment` at its mean. Create an interaction effect between `type` and the new `enrollment` variable. Re-run the model predicting `exempt`. Assume that `type` moderates the effect of enrollment in your interpretation of the interaction. Interpret the results on each coefficient. Create a plot to visualize the interaction.

```{r}
pbe <- pbe %>%
  mutate(enrollment_rct = as.vector(scale(enrollment, scale = FALSE))) # scale recenters based on mean and variance
# scale = FALSE that I just want to subtract the mean
# the as.vector call fixes an issue that came up with plotting
# interaction effects

pbe_lm2 <- lm(exempt ~ enrollment_rct + type + enrollment_rct:type, data = pbe)

summary(pbe_lm2)
```

```{r}
df <- ggeffects::ggpredict(pbe_lm2, terms = c("enrollment_rct", "type"))

ggplot(df, aes(x, predicted, colour = group)) + 
  geom_line()
```


## QUESTION 5

Let's log transform (using the natural log) the variable `enrollment` and call the new variable `log_enroll`. Estimate a model predicting `exempt` by `type` and `log_enrollment`. Interpret the coefficient on the log of enrollment. Does it make more sense to use enrollment or the log of enrollment as the predictor variable? Why?

```{r}
pbe <- pbe %>%
  mutate(log_enroll = log(enrollment))

pbe_lm3 <- lm(exempt ~ log_enroll + type, data = pbe)

summary(pbe_lm3)
```

## QUESTION 6

Create a binary variable to indicate high versus low `exempt` rates. For schools with exempt percentages equal to or greater than 33 percent, indicate them as "high", for all other schools indicate them as "low". Run a logistic regression predicting whether a school is high versus low, in other words, we want our model to predict schools falling into the high category. In your model use the predictors of `type` and `enrollment` (note: do not use `log_enroll` in this model). Interpret the coefficients on `type` and `enrollment` in terms of both log odds and odds. What is the probability of being a high `exempt` school if the school is private and has 100 students enrolled. 

```{r}
pbe <- pbe %>%
  mutate(exempt_cat = factor(ifelse(exempt >= 33, "high", "low")))

pbe_lm4 <- glm(exempt_cat ~ type + enrollment, data = pbe, family = "binomial")

summary(pbe_lm4)
```

# PART TWO

## QUESTION 7

A. Let's begin by exploring the data. How many years are there in the `concarry` data? How many observations per state?

```{r}
concarry %>%
  count(year) # 23 years

concarry %>%
  count(statename) # 23 obs per state
```

B. How many states had conceal carry laws (`shall`) in 1977 and how many had concealed carry laws in 1999?

```{r}
concarry %>%
  filter(year == 1977, shall == 1) %>%
  count(statename) # 4 states

concarry %>%
  filter(year == 1999, shall == 1) %>%
  count(statename) # 29 states
```

C. Create a plot tracking the violent crime rate (`vio`) over time for states that have ever adopted conceal carry laws (`shall`) and those that have never adopted the law. 

```{r}
concarry %>%
  group_by(stateid) %>%
  mutate(shall_never = factor(ifelse(sum(shall) > 0, 0, 1))) %>%
  group_by(shall_never, year) %>%
  summarize(vio = mean(vio)) %>%
  ggplot(aes(y = vio, x = year, col = shall_never)) +
  geom_line()
```


## QUESTION 8

Convert the violent crime rate (`vio`) into a logged variable (using the natural log), call it `log_vio`. This will be our dependent variable. Run a pooled regression of the data (i.e. standard OLS model as if this was cross-sectional data) predicting the log of violent crimes (`log_vio`) as a function of the presence of concealed carry laws (`shall`) and a set of dummy variables for year. Interpret the effect of `shall`. In general terms, what do the `year` dummy variables tell us about crime trends? In our current specification of the model, is the effect of `shall` the same for all years? Why or why not?

```{r}
concarry <- concarry %>%
  mutate(log_vio = log(vio),
         year = factor(year))

concarry_lm <- lm(log_vio ~ shall + year, data = concarry)

summary(concarry_lm)
```

